---
title: "Setup and Intro to SpatialExperiment"
output:
  xaringan::moon_reader:
    seal: true
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.retina = 2
)
```


# Installation

Everything you need for the workshop can be installed from `R` via `BiocManager`:

```{r install, eval=FALSE}
install.packages("BiocManager")
BiocManager::install()
BiocManager::install("bhuvad/CosMxSpatialAnalysisWorkshop")
```

Link to these slides:

```{r, echo=FALSE, out.width="30%"}
library(qrcode)
plot(qr_code("https://github.com/sagc-bioinformatics/spatialSlides2026"))
```
---

# The Bioconductor ecosystem


- A bioinformatics-focused package repository with relatively strict requirements, and an active community. 
- Many great R packages for single cell and spatial transcriptomics within the Bioconductor ecosystem. 
- Packages built upon common data structures mean Bioconductor packages are often interoperable.

---

# What is a SpatialExperiment?

Let's take advantage of the generally good documentation of the Bioconductor ecosystem and look at the vingette for SpatialExperiment to explain:

```{r, eval=FALSE}
browseVignettes("SpatialExperiment")
```

---

# Loading the example data

```{r}
library(SpatialExperiment)
# This may require BiocManager::install("VisiumIO")
example(SpatialExperiment, echo = FALSE)
spe
```

---

# Useful attributes

- store multiple transformations of assay data
- can put your image data in here (imgData)
- store spatial x/y coordinates
- store cell and gene metadata (rowData and colData)
- store embeddings and dimensionality reductions (reducedDims)
- probe or sequencing based spatial can both be analysed (as long as you collapse to cells or bins)
- other extentions (e.g. `SpatialFeatureExperiment`) can even integrate geometry


---


# SpatialExperiment's relationships

Most spatial packages in Bioconductor are built around `SpatialExperiment`.
`SpatialExperiment` actually extends from another Bioconductor package, `SingleCellExperiment`, which in turn extends `SummarizedExperiment`.

```{r}
is(spe, "SpatialExperiment")
is(spe, "SingleCellExperiment")
```

Many popular single cell packages (e.g. `scater`, `scran`) were written with `SingleCellExperiment` in mind.
But they can also accept classes that extend `SingleCellExperiment`, meaning we can use `SpatialExperiment` objects with these packages!

---


# Updating

Perhaps we want to log our counts:

```{r}
assays(spe)
```

First we'll have to ensure there are no zeros.
We can subset the SPE to filter zeros out:

```{r}
# subsetting returns a new object, we can overwrite the old one
spe <- spe[, colSums(counts(spe)) > 0]
```

We can then add a new assay update the object as our analysis progresses:

```{r}
library(scuttle)
spe <- logNormCounts(spe)
assays(spe)
```


---

# More on subsetting

Rows are genes, columns are cells.
Subsetting the SPE also subsets all slots and metadata in sync!

```{r subsetting, eval=FALSE}
# First 10 genes, all cells
spe[1:10, ]
# subset cells based on coldata column
spe[, spe$in_issue]

# subset by spatial coordinates
spe[, spatialCoords(spe)[, 1] > 2000]
```

---

# Plotting in Space

`plotSpatial` from `SpaNorm` returns a `ggplot` object:

```{r plot-region, eval=FALSE}
library(SpaNorm)

plotSpatial(spe, what = "expression", assay = "logcounts", colour = ENSMUSG00000051951)
```

---

# How do I build a SpatialExperiment?

**Importers exist for most platforms:**
```{r importers, eval=FALSE}
library(SpatialExperimentIO)
# from SpatialExperimentIO
spe <- readXeniumSXE("path/to/xenium/output/")
spe <- readCosMxSXE("path/to/cosmx/output/")
# built into SpatialExperiment
spe <- read10xVisium("path/to/spaceranger/output/")
```

Packages to import specific technologies can lag behind vendor updates,
but it's easy to construct one manually:

```{r build-spe, eval=FALSE}
library(Matrix)

spe <- SpatialExperiment(
  assays        = list(counts = counts_matrix),  # genes x cells
  rowData       = gene_metadata,    # data.frame, one row per gene
  colData       = cell_metadata,    # data.frame, one row per cell/spot
  spatialCoords = coords_matrix     # matrix with "x" and "y" columns
)
```

---

class: inverse, center, middle

# What's Next

Workshop: https://bhuvad.github.io/CosMxSpatialAnalysisWorkshop/articles/workflow_spatial_cosmx.html

```{r, echo=FALSE, out.width="30%"}
library(qrcode)
plot(qr_code("https://bhuvad.github.io/CosMxSpatialAnalysisWorkshop/articles/workflow_spatial_cosmx.html"))
```

